---
title: "Get started with samplesim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with samplesim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 7,
  out.width  = "100%",
  dpi        = 300
)
```

```{r 'setup', echo = FALSE}
library("samplesim")
library("ggplot2")
```

This package allows the investigation of the effect of sample size on estimates 
and precision of stable isotope mixing solutions calculated with the package 
[**siar**](https://cran.r-project.org/web/packages/siar/index.html) or the 
package [**MixSIAR**](https://cran.r-project.org/web/packages/MixSIAR/index.html). 
Samples sizes are modified assuming a normal distribution with a user defined 
mean and standard deviation. Samples of different sizes are created from this 
distribution, and mixing proportions are estimated for several replicates of 
each sample size using the function `siar::siarmcmcdirchletv4()` or the 
function `MixSIAR::run_model()`.



## Data description



Let's take a look at the data requirements. `samplesim` needs three types of data:

- the isotopic plasma values of the consumers;
- the isotopic plasma values of the preys (also called sources);
- the trophic discrimination factors.

Datasets provided in this package come from Killengreen _et al._ (2011)[^1]

[^1]: Killengreen S., Lecomte N., Ehrich D., Schott T., Yoccoz N.G. and Ims R.A. (2011) 
The importance of marine vs. human-induced subsidies in the maintenance of an expanding 
mesocarnivore in the arctic tundra. _Journal of Animal Ecology_, **80**, 1049-1060.

To use these datasets with `samplesim` we will use the functions 
`MixSIAR::load_mix_data()`, `MixSIAR::load_source_data()` and 
`MixSIAR::load_discr_data()` to import and format the data.



### Consumers data {#consumers}



This dataset contains the isotopic plasma values of the consumers over two 
(or more) isotopes. It must be structured in a data frame with two (or more) 
variables. Each column contains the isotopic plasma values of one isotope. In 
the example provided, this dataset contains two columns:

- `d13C`, describing the plasma values of the d13C isotope
- `d15N`, describing the plasma values of the d15N isotope

Each row of the data frame corresponds to one consumer. Let's take a look at the 
dataset available in the `samplesim` package by importing data with the 
`MixSIAR::load_mix_data()` function.



```{r, eval = TRUE, echo = TRUE}
## Get the location of the dataset ----
file_path <- system.file("extdata", "foxes_consumer.csv", package = "samplesim")

## Import isotopic plasma values of the consumers ----
consumer <- MixSIAR::load_mix_data(filename     = file_path,
                                   iso_names    = c("d13C", "d15N"),
                                   factors      = NULL,
                                   fac_random   = NULL,
                                   fac_nested   = NULL,
                                   cont_effects = NULL)
```

**Important:** currently `samplesim` is not develop to deal with random factors 
or continuous covariate.



<br />

Let's explore this dataset.

```{r, eval = TRUE, echo = TRUE}
## Get the class of the object ----
class(consumer)

## Name of the list elements ----
names(consumer)

## Print the ten first rows of the isotopic values ----
head(consumer$"data_iso", 10)

## Get the name of the isotopes ----
consumer$"iso_names"

## How many consumers are there in the dataset? ----
consumer$"N"
```



<br />

For further details on the function `MixSIAR::load_mix_data()` the reader is 
referred to the help page of this function.



### Sources data {#sources}



This dataset contains the isotopic plasma values of the sources over two 
(or more) isotopes. It must be structured in one of two following formats:

_**Format \#1**_

A data frame with repeated measurements of plasma isotope (columns) for each 
source individual (row). In the example provided in `samplesim`, this dataset 
contains three columns:

- `Source`: the name of the sources;
- `d13C`: the plasma values of the d13C isotope;
- `d15N`: the plasma values of the d15N isotope.

_**Format \#2**_

A data frame expressing isotopic plasma values as means and standard deviation. 
This dataset also contains the sample size of each source. In the example 
provided in `samplesim`, this dataset contains six columns:

- `Sources`: the name of the sources;
- `Meand13C`: the mean values of the d13C isotope;
- `SDd13C`: the standard deviation of the d13C isotope;
- `Meand15N`: the mean values of the d15N isotope;
- `SDd15N`: the standard deviation of the d15N isotope;
- `n`: the sample size of each source.

<br />

Let's take a look at the dataset available in the `samplesim` package by 
importing data with the function `MixSIAR::load_source_data()`.

```{r, eval = TRUE, echo = TRUE}
## Get the location of the dataset ----
file_path <- system.file("extdata", "foxes_sources.csv", package = "samplesim")

## Import mean isotopic plasma values of the sources (Format #1) ----
sources <- MixSIAR::load_source_data(filename       = file_path,
                                     source_factors = NULL,
                                     conc_dep       = FALSE,
                                     data_type      = "means",
                                     mix            = consumer)
```

**Note**: If your data follow the format \#2, replace the value of the argument 
`data_type` by `raw`.



<br />


Let's explore this dataset.

```{r, eval = TRUE, echo = TRUE}
## Get the class of the object ----
class(sources)

## Name of the list elements ----
names(sources)

## Print the mean values of isotopic plasma values ----
sources$"S_MU"

## Print the SD values of isotopic plasma values ----
sources$"S_SIG"

## Get the name of the sources ----
sources$"source_names"

## How many sources are there in the dataset? ----
sources$"n_array"
```

For further details on the function `MixSIAR::load_source_data()` the reader is 
referred to the help page of this function.



### Discrimation data {#discrim}



This dataset contains the trophic discrimination factor (TDF) of two (or more) 
isotopes for each source. It must be structured following the format \#2 of the 
sources dataset. In the example provided in `samplesim`, this dataset contains 
five columns:

- `Sources`: the name of the sources;
- `Meand13C`: the mean values of the TDF for the d13C isotope;
- `SDd13C`: the standard deviation of the TDF for the d13C isotope;
- `Meand15N`: the mean values of the TDF for the d15N isotope;
- `SDd15N`: the standard deviation of the TDF for the d15N isotope.

Let's take a look at the dataset available in the `samplesim` package by 
importing data with the function `MixSIAR::load_discr_data()`.

```{r, eval = TRUE, echo = TRUE}
## Get the location of the dataset ----
file_path <- system.file("extdata", "foxes_discrimination.csv", package = "samplesim")

## Import TDF values ----
discr <- MixSIAR::load_discr_data(filename = file_path, mix = consumer)
```

<br />

Let's explore this dataset.

```{r, eval = TRUE, echo = TRUE}
## Get the class of the object ----
class(discr)

## Name of the list elements ----
names(discr)

## Print the mean values of TDF ----
discr$"mu"

## Print the SD values of TDF ----
discr$"sig2"
```

For further details on the function `MixSIAR::load_discr_data()` the reader is 
referred to the help page of this function.



## Data visualization



The function `samplesim::plot_isospace()` represents consumer and sources data 
in the isotopic space. This function is an adaptation of `MixSIAR::plot_data()` 
function and add color to source data. If you want you can export plot in PDF 
and/or PNG format.


```{r, echo = TRUE, eval = TRUE, fig.width = 5, fig.height = 5}
samplesim::plot_isospace(mix = consumer, source = sources, discr = discr)
```

If you prefer you can also used the `MixSIAR::plot_data()`. 

If your data make sense, let's move forward!



## Running samplesim



The core function of the package is `samplesim::samplesim()`. It allows 
investigating the effect of sample size on estimates and precision of stable 
isotope mixing solutions. More specifically `samplesim::samplesim()` assesses 
the sensitivity of isotopes mixing models to variation in numbers of samples 
from source tissues. This tool can be used prior to full-blown studies in a 
similar manner than power analyses. It used the function 
`siar::siarmcmcdirichletv4()` developed by Andrew Parnell and available in the 
package **siar**. `samplesim` can also be used with the package **MixSIAR** 
developed by Brian Stock _et al._. User can choose to sample one particular 
source, or all the sources. User can also choose to modify consumer data.
Sample sizes are modified assuming a normal distribution with a user defined 
mean and standard deviation. Samples of different sizes are created from this 
distribution, and mixing proportions are estimated for several replicates of 
each sample size.



<br />



The general writing of `samplesim::samplesim()` is:

```{r, eval = FALSE, echo = TRUE}
samplesim::samplesim(package     = "siar",
                     mix         = consumer,
                     source      = sources,
                     discr       = discr,
                     type        = NULL,
                     nsamples    = NULL,
                     modify      = NULL,
                     nrep        = 100,
                     interval    = 90,
                     name        = NULL,
                     resid_err   = TRUE,
                     process_err = FALSE,
                     run         = "test",
                     alpha.prior = 1,
                     path        = ".")
```

with:

- `package`, the package name to be used to estimate mixing proportions. 
Must be one of `'siar'` or `'mixsiar'`.
- `mix`, the output returned by the `MixSIAR::load_mix_data()` function and 
containing consumer isotope values ([see Consumers data section](#consumers)).
- `source`, the output returned by the `MixSIAR::load_source_data()` function 
and containing mean and standard deviation isotope values of sources (and in 
some case raw values; [see Preys data section](#sources)).
- `discr`, the output returned by the `MixSIAR::load_discr_data()` function and 
containing TDF values ([see Discrimation data section](#discrim)).
- `type`, the type of analysis to be run. Must be one of `'one source'`, 
`'all sources'` or `'consumer'`.
- `nsamples`, a vector with the sample sizes to simulate.
- `modify`, the name of the source to modify (case sensitive). This argument 
has to be specified when the argument `type = 'one source'`. Otherwise 
it will be ignored.
- `nrep`, an integer specifying the number of replicates for each sample sizes. 
Default is `100`.
- `interval`, an integer indicating the width of credible interval to use for 
precision estimation. Default is `90`.
- `name`, a character string giving the name of the simulation. If `NULL` the 
simulation will be named by the time of the simulation. This name will be used 
to create a directory (in `path`) in which results will be stored.
- `resid_err`, a boolean indicating if residual error is included in the model. 
See `?MixSIAR::run_model` for further information. 
Only necessary if `package = 'mixsiar'`.
- `process_err`, a boolean indicating if process error is included in the model. 
See `?MixSIAR::run_model` for further information. 
Only necessary if `package = 'mixsiar'`.
- `run`, a string or a list specifying MCMC parameters. See `?MixSIAR::run_model` 
for further information. Only necessary if `package = 'mixsiar'`.
- `alpha.prior`, a numeric giving the Dirichlet prior on p.global. See 
`?MixSIAR::run_model` for further information. Only necessary 
if `package = 'mixsiar'`.
- `path`, the directory in which the directory `name` will be created.


Let's take an example. We will assess the impact of the sample size of 
one source: the source \#6 (_Voles_) with several sample sizes (from 2 to 500). 
The analysis will be repeated 999 times.

```{r, eval = FALSE, echo = TRUE, message = FALSE}
## samplesim run for one source ----
samplesim::samplesim(package     = "siar",
                     mix         = consumer,
                     source      = sources,
                     discr       = discr,
                     type        = "one source",
                     nsamples    = c(2:10, 15, 25, 50, 75, 100, 150, 250, 500),
                     modify      = "Voles",
                     nrep        = 999,
                     interval    = 90,
                     name        = "test_siar",
                     resid_err   = TRUE,
                     process_err = FALSE,
                     run         = "test",
                     alpha.prior = 1,
                     path        = ".")
```

This function does not return any object in the R console. Instead, results are 
stored in the directory `./test_siar/` and they will need to be imported for 
results visualization ([see below](#results)).

**Note 1:** if you want to estimate sample size impacts of all sources, you 
need to set `type = "all source"`. If you want to assess the impact of the 
sample size of consumers, you have to set `type = "consumer"`.

**Note 2:** the use of `samplesim` with the package `MixSIAR` requires the 
installation of the software **JAGS**. See the `MixSIAR` documentation for 
further details.



## Results extraction {#results}



The `samplesim::samplesim()` function has stored four objects results.

```{r, eval = FALSE, echo = TRUE}
## Objects created by samplesim ----
list.files("./test_siar")
```

```{r, eval = TRUE, echo = FALSE}
list.files("docs/test_siar")
```


- `intervals`, a four dimensions array with the upper and lower bounds of the 
credible interval for each sample size, replicate and source. First dimension 
represents lower and upper bounds; second dimension corresponds to the number 
of sources; third dimension is the number of replicates; and fourth dimension 
is the number of sample size.

- `widths`, a three dimensions array with the width (precision) of credible 
intervals for each source, each replicate and each sample size. First dimension 
corresponds to the number of replicates; second dimension is the number of 
sources; and third dimension represents the number of sample size.

- `medians`, a three dimensions array with the median (estimate) of credible 
intervals for each source, each replicate and each sample size. Dimensions are 
the same as for `widths` object.

- `datasets`, a four dimensions array with all resampled datasets.

A `logfile.txt` is also written and contains all parameters of the simulation. 
Here's what it looks like.

```{r, echo = FALSE, eval = TRUE}
cat(paste0(readLines("docs/test_siar/logfile.txt"), collapse = "\n"))
```



<br />



Now, let's take a look at the results by importing the `medians` dataset using 
the R function `readRDS()`.

```{r, echo = TRUE, eval = FALSE}
## Import medians of credible intervals ----
medians <- readRDS("./test_siar/medians.rds")
```

```{r, echo = FALSE, eval = TRUE}
# Import medians of credible intervals
medians <- readRDS("docs/test_siar/medians.rds")
```


```{r, echo = TRUE, eval = TRUE}
## Structure of the object ----
class(medians)

## Names of the dimensions ----
names(dimnames(medians))

## Names of the content of the second dimension ----
dimnames(medians)[[2]]

## Names of the content of the third dimension ----
dimnames(medians)[[3]]
```

These data are structured in a three dimensions array.

```{r, echo = TRUE, eval = TRUE}
## Extract results of the first replicate ----
medians[1, , ]

## Compute mean over replicates ----
apply(medians, 3:2, mean)
```

The structure of the `widths` dataset is the same as for `medians`.



<br />



If you are uncomfortable to deal with three dimensions arrays, you can use the 
function `samplesim::get_output()`. This function converts these two arrays 
(i.e. `medians` and `widths`) into an unique data frame. For instance:

```{r, echo = TRUE, eval = FALSE}
## Import medians and widths of credible intervals ----
datas <- samplesim::get_output("test_siar")
```

```{r, echo = FALSE, eval = TRUE}
datas <- samplesim::get_output("test_siar", path = "docs")
```

```{r, echo = TRUE, eval = TRUE}
## Structure of the data frame ----
str(datas)

## Print the first ten and last ten rows ----
rbind(head(datas, 10), tail(datas, 10))
```

If you want to select only the `widths` data,

```{r, echo = TRUE, eval = TRUE}
## Extract widths of credible intervals ----
widths <- datas[datas$type == "Width of credible intervals", ]

## Check ----
rbind(head(widths, 10), tail(widths, 10))
```

The argument `change` of the function `samplesim::get_output()` allows to 
compute the percentage of change of medians and widths respectively, based on a 
reference value of sample size (argument `reference`; default is the minimum 
sample size). For instance,

```{r, echo = TRUE, eval = FALSE}
## Import medians and widths of credible intervals expressed as percentage of change ----
datas <- samplesim::get_output(name = "test_siar", change = TRUE, reference = 2)
```

```{r, echo = FALSE, eval = TRUE}
datas <- samplesim::get_output(name = "test_siar", change = TRUE, reference = 2, path = "docs")
```

```{r, echo = TRUE, eval = TRUE}
## Structure of the data frame ----
str(datas)

## Print the first ten and last ten rows ----
rbind(head(datas, 10), tail(datas, 10))
```

Here we notice that the column replicate is omitted and percentages of change 
are aggregated over replicates.



## Results visualization



The `samplesim` package also offers the possibility of visualizing the effect 
of sample size on the width of the credible interval and on the median of the 
posterior distribution of the mixing models. This is possible with the function 
`samplesim::plot_samplesim()`.


```{r, echo = TRUE, eval = FALSE, fig.width = 8.5, fig.height = 5.5}
## Visualize results ----
samplesim::plot_samplesim(name = "test_siar")
```

```{r, echo = FALSE, eval = TRUE, fig.width = 8.5, fig.height = 5.5}
# Visualize results
samplesim::plot_samplesim(name = "test_siar", path = "docs")
```



<br />



If you prefer you can represent the percentages of change:

```{r, echo = TRUE, eval = FALSE, fig.width = 8.5, fig.height = 5.5}
# Visualize results expressed as percentages of change
samplesim::plot_samplesim(name = "test_siar", change = TRUE, reference = 2)
```

```{r, echo = FALSE, eval = TRUE, fig.width = 8.5, fig.height = 5.5}
samplesim::plot_samplesim(name = "test_siar", change = TRUE, reference = 2, path = "docs")
```


<br />


Finally, if you are not satisfied with the quality of the graph, you can 
customize it by 1) importing data and 2) programming your own graph. Let's take 
an example with the 
[**ggplot2**](https://cran.r-project.org/web/packages/ggplot2/index.html) 
package (used in `samplesim`).

```{r, echo = TRUE, eval = FALSE}
## Import medians and widths of credible intervals ----
datas <- get_output("test_siar")
```

```{r, echo = FALSE, eval = TRUE}
datas <- get_output("test_siar", path = "docs")
```

```{r, echo = TRUE, eval = TRUE, fig.width = 8.5, fig.height = 5.5}
ggplot(aes(x = size, y = value), data = datas) +
  geom_boxplot(aes(fill = source), width = 0.8, outlier.shape = NA) +
  labs(x = "Sample size", y = "Values", fill = "Sources") +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  facet_grid(. ~ type)
```



<br />



Or your can only represent the medians values. For instance (with percentages of 
change):


```{r, echo = TRUE, eval = FALSE}
## Import medians and widths of credible intervals ----
datas <- samplesim::get_output(name = "test_siar", change = TRUE)
```

```{r, echo = FALSE, eval = TRUE}
datas <- samplesim::get_output(name = "test_siar", change = TRUE, path = "docs")
```


```{r, echo = TRUE, eval = TRUE, fig.width = 8.5, fig.height = 5.5}
## Select only medians of credible intervals ----
medians <- datas[datas$"type" == "Median of posterior distribution", ]

ggplot(aes(x = size, y = value, group = source), data = medians) +
  geom_point(aes(color = source)) +
  geom_line(aes(color = source)) +
  labs(x = "Sample size", y = "Change in medians (%)", color = "Sources") +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())
```


## Reference {-}

Lecomte N., Ehrich D., Casajus N., Berteaux D., Giroux M.-A. and Yoccoz N.G. 
How many is enough? An R package for evaluating the effect of sample size on 
estimates and precision of stable isotope mixing solutions. 
Submitted to _Methods in Ecology and Evolution_.
